### 2015-07-26 ms-thesis/ver1/kalman/ms-thesis-kalman.tex
-

\documentclass[a4paper,11pt]{article}

\usepackage[latin5]{inputenc}
\usepackage{amsmath}
\usepackage{examplep}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphics}

\title{Kalman Filters: Derivation}

\author{Burak Bayramlý}

\begin{document}

\maketitle

@@

@@

\begin{thebibliography}{99}

\bibitem{rabiner} Rabiner L. R. , 
  {\em A Tutorial on Hidden Markov Models and Selected Applications in Speech Recognition}, 
  Proceedings of IEEE vol. 77, no. 2, pp. 257-286, 
  1989.

\bibitem{ghahramani} Roweis S. and Z. Ghahramani,
  {\em  A Unifying Review of Linear Gaussian Models },
   Neural Computation 11(2):305--345, 
   1999.

\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
  {\em Parameter Estimation for Linear Dynamical Systems },
   Technical Report CRG-TR-96-2  
   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
   Department of Computer Science, University of Toronto, 
   1996.

\bibitem{ghahramani3} Ghahramani Z., H. E. Hinton,
  {\em Switching State Space Models},
  Technical Report CRG-TR-96-3, Dept. Comp. Sci., Univ. Toronto, 
  1996.

\bibitem{shumway} Shumway R., H. S. Stoffer
  {\em Time series analysis and its applications 2nd Edition},
   New York, Springer, (Springer texts in statistics),
   2000.

\bibitem{jordan} Jordan M. I. , C. Bishop
  {\em An Introduction to Graphical Models},
   Not yet published, 
   2000.

\bibitem{kalman} Kalman R. E.,
  {\em A New Approach to Linear Filtering and Prediction Problems},
  Transactions of the ASME-Journal of Basic Engineering, 82 (Series D): 35-45, 
  1960.

\bibitem{kalman2} Kalman, R.E. and R.S. Bucy, 
  {\em New results in filtering and prediction theory},
  Trans. ASME J. Basic Eng., 83, 95-108,
  1961.

\bibitem{welling} Welling, M.,
  {\em The Kalman Filter - Lecture Tutorial},
   California Institute of Technology,
   2008.

\bibitem{lall} Lall, S.,
  {\em Modern Control 2 Lecture Notes},
   Stanford University,
   2006.

\bibitem{tsay} Tsay, R. S.,
  {\em Time Series and Forecasting: Brief History and Future Research,},
  Journal of the American Statistical Association, Vol. 95, No. 450., pp. 638-643.,
  Jun 2000.

\bibitem{shi} Shi S. and A. S. Weigend., {\em Taking Time Seriously: Hidden
  Markov Experts Applied to Financial Engineering}. In Proceedings of the
  IEEE/IAFE 1997 Conference on Computational Intelligence for Financial
  Engineering , pages 244-252. IEEE, 
  1997.

\bibitem{mandelbrot} Mandelbrot B. B., 
  {\em The variation of certain speculative prices, Journal of Business},
  XXXVI (1963), pp. 392-417, 
  1963.

\end{thebibliography}

\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{KALMAN FILTERS}

@@

\begin{figure}[!hbp]
\center{
  \scalebox{0.6}{
  \includegraphics{./kalman/kf.JPG}
  }
}
\caption{Hidden and observed states of a KF}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Smoothing}

Filtering involves applying a new measurement $y_{t+1}$ to calculate
$\hat{x}_{t+1}^{t+1}$ and $P_{t+1}^{t+1}$ which is an excellent tool for an
online algorithm. However for stock price prediction, we will have all data we
need already at hand. As opposed to a tracking application where data cannot be
known in advance, we will have our data available, as measurements
$y_1,...,y_T$, in the form of stock prices, to train a Kalman filter based
solution, all at once. Hence we need an equation for $x_{t}'$. For this we need
RTS (Rauch-Tung-Striebel) Smoother.

Based on \eqref{jointxy}, we can write the covariance of joint distribution
$(x_t,x_{t+1})$. In \eqref{jointxy} the multiplication was with $C$, here it's
$A$. Hence,
\begin{eqnarray*}
\left[ \begin{array}{cc}
      P_{t}^t & P_{t}^tA' \\
      AP_{t}^t & AP_{t}^tA'+R
      \end{array} \right]
\end{eqnarray*}
Combining the previous equations together, we need to reach to a statement for
$x_{t}'$. We can start with $\mathbf{E}[x_t|x_{t+1},y_0,...,y_t)]$.
\begin{eqnarray*}
\mu_{1|2} &=& \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}(x_2 - \mu_2)
\end{eqnarray*}
which becomes
\begin{eqnarray*}
\mathbf{E}[x_t|x_{t+1},y_0,...,y_t)] = \hat{x}_{t}^t+P_{t}^tA'(P_{t+1}^t)^{-1}(x_{t+1}-\hat{x}_{t+1}^t)
\end{eqnarray*}
Define $L_t \triangleq P_{t}^tA'(P_{t+1}^t)^{-1}$. Then,
\begin{eqnarray*}
\mathbf{E}[x_t|x_{t+1},y_0,...,y_t)] =
\hat{x}_{t}^t+L_t(x_{t+1}-\hat{x}_{t+1}^t)
\end{eqnarray*}
How about $Var[x_t|x_{t+1},y_0,...,y_t]$? If we remember the joint covariance
for $(x_t,x_{t+1})$
\begin{eqnarray*}
 \left[ \begin{array}{cc} 
      P_{t}^t & P_{t}^tA' \\
      AP_{t}^t & AP_{t}^tA'+R
      \end{array} \right] && \\
\end{eqnarray*}
Then
\begin{eqnarray*}
\Sigma_{1|2} &=& \Sigma_{11}-\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}
\end{eqnarray*}
becomes
\begin{eqnarray*}
Var[x_t|x_{t+1},y_0,...,y_t] &=& P_{t}^t - P_{t}^tA'(P_{t+1}^t)^{-1}AP_{t}^t
\end{eqnarray*}
Why?
\begin{eqnarray*}
\left[ \begin{array}{cc}
      P_{t}^t & P_{t}^tA' \\
      AP_{t}^t & AP_{t}^tA'+R
      \end{array} \right]
=      
\left[ \begin{array}{cc}
      P_{t}^t & P_{t}^tA' \\
      AP_{t}^t & P_{t+1}^t
      \end{array} \right]
\end{eqnarray*}
If we try to substitute $L_t \triangleq P_{t}^tA'(P_{t+1}^t)^{-1}$ in to the
equation:
\begin{eqnarray*}
Var[x_t|x_{t+1},y_0,...,y_t] &=& P_{t}^t - P_{t}^t A'(P_{t+1}^t)^{-1}AP_{t}^t \\
&=& P_{t}^t - P_{t}^tA'(P_{t+1}^t)^{-1} (P_{t+1}^t(P_{t+1}^t)^{-1}) AP_{t}^t\\
&=& P_{t}^t - (P_{t}^tA'(P_{t+1}^t)^{-1}) P_{t+1}^t ((P_{t+1}^t)^{-1} AP_{t}^t)\\
&=& P_{t}^t - L_t  P_{t+1}^t L_t'
\end{eqnarray*}
Due to conditional independence, when we condition on $x_{t+1}$, we make $x_t$
independent of future observations $y_{t+1},...,y_T$. Hence we can simply say;
\begin{eqnarray*}
E[x_t|x_{t+1},y_0,...,y_T]  &=& E[x_t|x_{t+1},y_0,...,y_t]\\
&=& \hat{x}_{t}^t +L_t(x_{t+1}-\hat{x}_{t+1}^t) \\
Var[x_t|x_{t+1},y_0,...,y_T]  &=& Var[x_t|x_{t+1},y_0,...,y_t] \\
&=& P_{t}^t - L_t  P_{t+1}^t L_t'
\end{eqnarray*}
We are almost done. If we can drop the $x_{t+1}$ from LHS of both equations, we
can have our smoothing equations. Due to the so-called Tower Property:
\begin{eqnarray*}
\mathbf{E}[X|Z] &=& \mathbf{E}[\mathbf{E}[X|Y,Z]|Z]\\
Var[X|Z] &=& Var[\mathbf{E}[X|Y,Z]|Z] + \mathbf{E}[Var[X|Y,Z]|Z]
\end{eqnarray*}
we can drop $x_{t+1}$ from LHS.
\begin{eqnarray}
\hat{x}_{t}' & \triangleq & \mathbf{E}[x_t|y_0,...,y_T] \nonumber\\
& = & \mathbf{E}[\mathbf{E}[x_t|x_{t+1},y_0,...,y_T]|y_0,...,y_T] \nonumber\\
& = & \mathbf{E}[\hat{x}_{t}^t +L_t(x_{t+1}-\hat{x}_{t+1}^t)] \nonumber\\
& = & \hat{x}_{t}^t +L_t(\hat{x}_{t+1}' -\hat{x}_{t+1}^t) \label{xsmooth}
\end{eqnarray}
Using similar algebra, we get
\begin{eqnarray*}
P_{t}'  = P_{t}^t + L_t(P_{t+1}'-P_{t+1}^t)L_t'
\end{eqnarray*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Training}

In order to train a Kalman Filter based predictor, we need to estimate
parameters $A,C,Q,R,\pi$. These parameters define a KF.

Making use of Markov property, we write our joint probability as;

\begin{eqnarray*}
p(x_0,..,x_t,y_0,..,y_t) =
p(x_0)\prod_{t=1}^{T-1}p(x_{t+1}|x_t)\prod_{t=0}^{T-1}p(y_t|x_t)
\end{eqnarray*}

\begin{eqnarray*}
p(y_{t}|x_t) &=& exp
\bigg\{
-\frac{1}{2}(y_{t}-Cx_t) ' R^{-1}(y_{t}-Cx_t)
\bigg\}
(2\pi)^{-\frac{p}{2}}|R|^{-\frac{1}{2}}\\
p(x_{t+1}|x_t) &=& exp
\bigg\{
-\frac{1}{2}(x_{t+1}-Ax_t) ' Q^{-1}(x_{t+1}-Ax_t)
\bigg\} \\
&& (2\pi)^{-\frac{k}{2}}|Q|^{-\frac{1}{2}}\\
p(x_{1}) &=& exp
\bigg\{
-\frac{1}{2}(x_{1}-\pi) ' \Sigma_{\pi}^{-1}(x_{1}-\pi)
\bigg\}
(2\pi)^{-\frac{k}{2}}|\Sigma_{\pi}|^{-\frac{1}{2}}
\end{eqnarray*}
We take log
\begin{eqnarray*}
\log p(x,y) &=& \\
&& -\sum_{t=0}^{T-1}(-\frac{1}{2}(y_{t}-Cx_t) ' R^{-1}(y_{t}-Cx_t)) - \frac{T}{2}\log|R|\\
&& -\sum_{t=1}^{T-1}(-\frac{1}{2}(x_{t+1}-Ax_t) ' Q^{-1}(x_{t+1}-Ax_t))\\
&& - \frac{T-1}{2}\log|Q|\\
&& -\frac{1}{2}(x_{1}-\pi) ' \Sigma_{\pi}^{-1}(x_{1}-\pi) - \frac{T(p+k)}{2}\log|\Sigma_{\pi}|
\end{eqnarray*}

In order to estimate parameters, we take derivative of expected log likelihood,
set to zero and solve

\begin{eqnarray*}
Q &=& \mathbf{E}[\log p(x,y|y)] \\
\frac{\partial Q}{\partial C} &=& -\sum_{t=1}^{T}R^{-1}y_t \hat{x}_{t}' +
\sum_{t=1}' R^{-1} C P_t = 0
\end{eqnarray*}

\begin{eqnarray*}
C^{new} &=& \bigg( \sum_{t=1}' y_t \hat{x}_t' \bigg) \bigg( P_{t}' \bigg) ^{-1}
\end{eqnarray*}

Using similar technique, we can calculate other constants as;
\begin{eqnarray*}
R^{new} &=& \frac{1}{T}\sum_{t=1}^T \big( y_ty_t' - C^{new}\hat{x}_ty_t' \big) \\
A^{new} &=& \big( \sum_{t=2}^T P_{t,t-1})(\sum_{t=2}^TP_{t-1} \big) ^{-1} \\
Q^{new} &=& \frac{1}{T-1} \big( \sum_{t=2}^T P_t - A^{new}\sum_{t=2}^TP_{t-1,t} \big)\\
\pi_1^{new} &=& \hat{x}_1
\end{eqnarray*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage}

While using Kalman Filters for analysis and prediction purposes, one needs to be
careful about starting condition of the training algorithm. Because of the
nature of the training algorithm, the parameters $\mu_i, \sigma_i,A,C,Q,R,P$ are
all used for likelihood calculation/estimation and are {\em themselves} targets
for estimation. The ``chicken and egg'' nature of the EM training process
required us to start somewhere, and that somewhere is picking appropiate
parameters for the parameters mentioned above.

We had to determine said values by trial and error, as part of the process where
we were trying to determine meta parameters such as number of iterations
required for MC process, the lowest likelihood value threshold (which was used
to determine when to stop in the training process - if likelihood difference
fell below this threshold value, we would conclude that training was complete).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithm}

The algorithm that applies forward-backward algorithm and learning through EM is
described below \cite{ghahramani}.

\begin{figure}[!hbp]
\center{
\begin{tabbing}
{\tt LDS}\={\tt Inf}\={\tt erence}$(Y,A,C,Q,R,x,P)$\\
\> {\tt for } $t = 1$ {\tt to } $T$ \% forward recursion \\
\> \> $x_t^{t-1} = Ax_{t-1}^{t-1}$\\
\> \> $P_t^{t-1} = AP_{t-1}^{t-1}A' + Q$ {\tt if } $t>1$\\
\> \> $K_t = P_t^{t-1}C'(CP_t^{t-1}C'+R)^{-1}$\\
\> \> $x_t^t = x_t^{t-1} + K_t(y_t - Cx_t^{t-1})$\\
\> \> $P_t^t = P_t^{t-1} - K_tCP_t^{t-1}$\\
\> {\tt end}\\
\> {\tt initialize } $\hat{P}_{T,T-1}=(I-K_TC)AP_{T-1}^{T-1}$\\
\> \> {\tt for } $t=T$ {\tt to } $2$ \% backward recursion \\
\> \> $J_{t-1} = P_{t-1}^{t-1}A'(P_t^{t-1})^{-1}$\\
\> \> $\hat{x}_{t-1} = x_{t-1}^{t-1} + J_{t-1}(\hat{x}_{t} - Ax_{t-1}^{t-1})$\\
\> \> $\hat{P}_{t-1} = P_{t-1}^{t-1} + J_{t-1}(\hat{P}_t - P_t^{t-1})J_{t-1}'$\\
\> \> $\hat{P}_{t,t-1} = P_t^t J_{t-1}' + J_t(\hat{P}_{t+1,t} - AV_t^t)J_{t-1}'$
   {\tt if } $t<T$\\
\> {\tt end}\\
\> {\tt return }$\hat{x}_t, \hat{P}_t, \hat{P}_{t,t-1}$ {\tt for all } $t$
\end{tabbing}
}
\caption{LDS Inference Algorithm}
\vspace{0.6cm}
\end{figure}

\begin{figure}[!hbp]
\center{
\begin{tabbing}
{\tt LDS}\={\tt Learn}\=$(Y,k,$\=$\epsilon)$\\ \> {\tt initialize } $A, C, Q, R,
x_1^0, P_1^0$\\ \> $\alpha = \sum_t y_ty_t'$\\ \> {\tt while change in
  likelihood >} $\epsilon$\\ \> \> {\tt LDSInference}$(Y,A,C,Q,R,x_1^0,P_1^0)$
\% E-step\\ \> \> {\tt initialize } $\delta = 0, \gamma = 0, \sigma = 0$\\ \> \>
   {\tt for } $t = 1$ {\tt to } $T$\\ \> \> \> $\delta = \delta +
   y_t\hat{x}_t$\\ \> \> \> $\gamma = \gamma + \hat{x}_t\hat{x}_t' +
   \hat{P}_t$\\ \> \> \> $\beta = \beta + \hat{x}_t\hat{x}_{t-1}' + P_{t,t-1}$
       {\tt if } $t>1$\\ \> \> {\tt end}\\ \> \> $\gamma_1 = \gamma -
       \hat{x}_T\hat{x}_T' - \hat{P}_T$\\ \> \> $\gamma_2 = \gamma -
       \hat{x}_1\hat{x}_1' - \hat{P}_1$\\ \> \> \% M-step\\ \> \> $C = \delta
       \gamma^{-1}$\\ \> \> $R = (\gamma - C\delta')/T$\\ \> \> $A =
       \beta\gamma_1^{-1}$\\ \> \> $Q = (\gamma_2 - A\beta') / (T-1)$\\ \> \>
       $x_1^0 = \hat{x}_1$\\ \> \> $P_1^0 = \hat{P}_1$\\ \> {\tt end}\\ \> {\tt
         return } $A, C, Q, R, x_1^0, P_1^0$
\end{tabbing}
}
\caption{LDS Learning Algorithm}
\vspace{0.6cm}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Prediction with Kalman Filters}

Generating future values with a KF model is somewhat easier than generating
random values from an HMM. We know state space model assigns one hidden Gaussian
distribution per data point, which itself is modeled by another Gaussian. In
comparison, the number of hidden states of an HMM is specified by the modeler
and is usually less than the number of data points used for training.

The fact that we have as many hidden/observed random variables as data points
makes the following more intuitive: After the training phase of the KF which
includes forward recursion and smoothing, we have estimates for the mean and
variance for all Gaussian distributions. Another knowledge we have at hand is
the basic time transition equation for KFs - if we have an estimate at $t$ for
hidden state, we know how to propagate those estimates to the future at
$t+1$. Therefore, once training is complete, we can simply fast forward to the
emission Gaussian at time $T$, and using propagation equation generate a new
observation Gaussian at time $T+1$ which we use to generate a random number that
represents ``generatively'' an output at time $T+1$.

One point we have to be careful with is propagating the emission Gaussian, not
simply the hidden Gaussian. In the literature the basic time transition equation
is for the latter not the former, hence we need to expand the equation for $y_t$
by taking time transition into account:

Given,
\begin{eqnarray}
x_{t+1} = Ax_t + v_t \\
y_t = Cx_t + w_t
\end{eqnarray}
where $v_0,v_1,...$ is white Gaussian noise with $\mathbf{E}v_t = 0$ and $cov \:
v_t = Q$ and $w_0,w_1,...$ are also white Gaussian noise with $\mathbf{E}w_t =
0$ and $cov \: w_t = R$, we can determine $y_{t+1}$ as
\begin{eqnarray}
y_{t+1} &=& Cx_{t+1} + w_{t+1}\\
&=& C(Ax_t + v_t) + w_{t+1} \\
&=& CAx_t + Cv_t + w_{t+1}
\end{eqnarray}
which gives us the distribution of the emission Gaussian $y_{t+1}$. The
conditional expectation and variance of this distribution given $y_{1,..,t}$ is
\begin{eqnarray}
y_{t+1} &=& Cx_{t+1} + w_{t+1}\\
&=& C(Ax_t + v_t) + w_{t+1} \\
&=& CAx_t + Cv_t + w_{t+1} \\
\hat{y}_{t+1|t} &=& CA\hat{x}_{t|t} \\
cov(y_{t+1}|y_{0,...,t}) &=& CA P_{t|t}A'C' + CQC' + R
\end{eqnarray}
We can apply this recursive formula to have our KF generate $M$ many random
observations starting at $T$ up to $T+N$. The algorithm is

\begin{figure}[!hbp]
\center{
\begin{tabbing}
{\tt ob}\={\tt ser}\={\tt vation}\={\tt Genera}\={\tt tion}$(A,C,R,Q,\mu,\Sigma,T,N)$\\
\> $mu_x = mu[T-1];$\\
\> $sigma_x = sigma[T-1];$\\
\> {\tt for} $i = 1$ {\tt to } $N$\\
\> \> $mu_y = C A mu_x;$\\
\> \> $sigma_y = CA sigma_x A'C' + CQC' + R;$\\
\> \> $o[i] = $ {\tt gaussianNumberGeneration}$(mu_y, sigma_y);$\\
\> \> $mu_x = A mu_x$\\
\> \> $sigma_x = A  sigma_x A + Q$\\
\> {\tt end}\\
\> {\tt return} $o;$
\end{tabbing}
}
\caption{Generating Observations from an LDS Algorithm}
\vspace{0.6cm}
\end{figure}

### 2015-07-26 ms-thesis/ver2/biblio/bib.tex
-

\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
  {\em Parameter Estimation for Linear Dynamical Systems },
   Technical Report CRG-TR-96-2  
   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
   Department of Computer Science, University of Toronto, 
   1996.

\bibitem{alpaydin} Alpaydın E., 
  {\em Introduction to Machine Learning},
  MIT Press, 
  2004.

@@

\bibitem{nabney} Azzouzi M.  and Ian T. Nabney,
  {\em Modelling Financial Time Series with Switching State Space Models},
  \PVerb!citeseer.ist.psu.edu/267552.html!,
  1999.

>>>>

\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
  {\em Parameter Estimation for Linear Dynamical Systems },
   Technical Report CRG-TR-96-2  
   \url{ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz},
   Department of Computer Science, University of Toronto, 
   1996.

\bibitem{alpaydin} Alpaydin E., 
  {\em Introduction to Machine Learning},
  MIT Press, 
  2004.

@@

\bibitem{nabney} Azzouzi M.  and Ian T. Nabney,
  {\em Modelling Financial Time Series with Switching State Space Models},
  \url{citeseer.ist.psu.edu/267552.html},
  1999.

### 2015-07-26 ms-thesis/ver2/cover/ms-thesis-cover.tex
-

\title{FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN MARKOV MODELS}
\turkcebaslik{HMM VE KALMAN FİLTRELERİ KULLANARAK FİNANSAL ZAMAN SERİSİ TAHMİNİ}

@@

\author{Burak Bayramlı}

@@

\supervisor{Prof. Fikret Gürgen}

\examineri{Prof. Levent Akın}

\examinerii{Prof. Ethem Alpaydın}

>>>>

\title{FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN
  MARKOV MODELS \\ \\ HMM VE KALMAN FILTRELERI KULLANARAK FINANSAL ZAMAN SERISI TAHMINI}

@@

\author{Burak Bayramli}

@@

\supervisor{Prof. Fikret Gurgen}

\examineri{Prof. Levent Akin}

\examinerii{Prof. Ethem Alpaydin}

### 2015-07-26 ms-thesis/ver2/hmm/ms-thesis-hmm.tex
-

Hidden Markov Models are based on over hundred years old mathematical construct
called Markov Chains (MC). An MC can be in one of distinct $N$ states and the
transition between those states occur according to predefined transition
probabilities at discrete intervals. For the simplest, first-order Markov Chain,
these probabilities can be captured in a \PVerb!NxN!  matrix. The fact that
present state is enough to represent a system and is the only factor that
determines the future (through the transition table) is also called as the
system having the ``Markov property''. This also means that given present state,
the past and future are independent of eachother. Mathematically, a Markov Chain
is a sequence of random variables $X_1,X_2,X_3,...$, the state of the system at
time $t$ being $x_t$, then
\begin{eqnarray*}
P(X_{t+1} = x | X_t = x_t,...,X_1=x_1) = P(X_{t+1} = x | X_t = x_t)
\end{eqnarray*}
A Markov Chain could also be second order or third order and so on - for example
for second order Markov Chains, the transition table is \PVerb!2N x N! which
necessitates both present and {\em previous} step to obtain a transition probability
distribution for the next state.

@@

\begin{figure}[!hbp]
\center{
  \scalebox{0.8}{
  \includegraphics{./hmm/hmm-intro.jpg}
  }
}
\caption{Hidden and observed states of an HMM}
\vspace{0.6cm}
\end{figure}

@@

\begin{figure}[!hbp]
\center{
  \scalebox{0.8}{
  \includegraphics{./hmm/hmm-dep.jpg}
  }
}
\caption{Dependency between nodes of an HMM}
\vspace{0.6cm}
\end{figure}

@@

\begin{eqnarray*}
p(q_t|y) &=& \frac{p(y|q_t) p(q_t)}{p(y)}\\
&=& \frac{p(y_0...y_t|q_t)p(y_{t+1}...y_T|q_t) p(q_t)}{p(y)}\\
&=& \frac{p(y_0...y_t,q_t)p(y_{t+1}...y_T|q_t)}{p(y)}\\
p(q_t|y) &=& \frac{\alpha(q_t)\beta(q_t)}{p(y)}
\end{eqnarray*}
where
\begin{eqnarray*}
\alpha(q_t) & \triangleq & p(y_0...y_t,q_t) \\
\beta(q_t) & \triangleq & p(y_{t+1}...y_T|q_t)
\end{eqnarray*}
Obtain $p(y)$ by applying $\sum_{q_t}$ on both sides
\begin{eqnarray}
\sum_{q_t}p(q_t|y)  &=& \sum_{q_t} \frac{\alpha(q_t)\beta(q_t)}{p(y)} \nonumber\\
1  &=& \sum_{q_t} \frac{\alpha(q_t)\beta(q_t)}{p(y)} \nonumber\\
p(y) &=& \sum_{q_t} \alpha(q_t)\beta(q_t) \label{py}
\end{eqnarray}

@@

In order to find the hidden state at day $T$, we can simply run the Viterbi
algorithm that gives us the most likely hidden path for the given data and given
HMM. Then, we simply pick the last element from this list, roll the dice and
apply transition once, and this gives us our first prediction for day $T+1$. The
rest will be the same as generating data from a regular HMM - we keep following
the transition table. The pseucode for the function
\PVerb!observationGeneration!  looks as follows:

@@

\begin{figure}[ht]
\center{
\begin{tabbing}
{\tt next}\={\tt Sta}\={\tt te}$(stat$\=$e,A,\pi,C,R);$\\ 
\> $randVal = ${\tt generateUniform(0,1);}\\
\> $total = 0;$\\
\> $done = false;$\\
\> $total = A[currentState][0];$\\
\> {\tt for } $j = 1$ {\tt to } $numStates$\\
\> \> {\tt if} $done = true$ {\tt break};\\
\> \> {\tt if} $(randVal > total$ {\tt and } \\
\> \> \> $randVal \leqslant (total + A[currentState][j]))$\\
\> \> \> $currentState = j;$\\
\> \> \> $done = true;$\\
\> \> {\tt end}\\
\> \> $total = total+ A[currentState][j]$;\\
\> {\tt end}\\
\> {\tt return} $currentState$;
\end{tabbing}
}
\caption{State Transition Algorithm}
\vspace{0.6cm}
\end{figure}

>>>>

Hidden Markov Models are based on over hundred years old mathematical construct
called Markov Chains (MC). An MC can be in one of distinct $N$ states and the
transition between those states occur according to predefined transition
probabilities at discrete intervals. For the simplest, first-order Markov Chain,
these probabilities can be captured in a \verb!NxN!  matrix. The fact that
present state is enough to represent a system and is the only factor that
determines the future (through the transition table) is also called as the
system having the ``Markov property''. This also means that given present state,
the past and future are independent of eachother. Mathematically, a Markov Chain
is a sequence of random variables $X_1,X_2,X_3,...$, the state of the system at
time $t$ being $x_t$, then
\begin{eqnarray*}
P(X_{t+1} = x | X_t = x_t,...,X_1=x_1) = P(X_{t+1} = x | X_t = x_t)
\end{eqnarray*}
A Markov Chain could also be second order or third order and so on - for example
for second order Markov Chains, the transition table is \verb!2N x N! which
necessitates both present and {\em previous} step to obtain a transition probability
distribution for the next state.

@@

\begin{figure}[!hbp]
\center{
  \scalebox{0.8}{
  \includegraphics{./hmm/hmm-intro.JPG}
  }
}
\caption{Hidden and observed states of an HMM}
\vspace{0.6cm}
\end{figure}

@@

\begin{figure}[!hbp]
\center{
  \scalebox{0.8}{
  \includegraphics{./hmm/hmm-dep.JPG}
  }
}
\caption{Dependency between nodes of an HMM}
\vspace{0.6cm}
\end{figure}

@@

\begin{eqnarray*}
p(q_t|y) &=& \frac{p(y|q_t) p(q_t)}{p(y)}\\
&=& \frac{p(y_0...y_t|q_t)p(y_{t+1}...y_T|q_t) p(q_t)}{p(y)}\\
&=& \frac{p(y_0...y_t,q_t)p(y_{t+1}...y_T|q_t)}{p(y)}\\
p(q_t|y) &=& \frac{\alpha(q_t)\beta(q_t)}{p(y)}
\end{eqnarray*}
where
\begin{eqnarray*}
\alpha(q_t) & \equiv & p(y_0...y_t,q_t) \\
\beta(q_t) & \equiv & p(y_{t+1}...y_T|q_t)
\end{eqnarray*}
Obtain $p(y)$ by applying $\sum_{q_t}$ on both sides
\begin{eqnarray}
\sum_{q_t}p(q_t|y)  &=& \sum_{q_t} \frac{\alpha(q_t)\beta(q_t)}{p(y)} \nonumber\\
1  &=& \sum_{q_t} \frac{\alpha(q_t)\beta(q_t)}{p(y)} \nonumber\\
p(y) &=& \sum_{q_t} \alpha(q_t)\beta(q_t) \label{py}
\end{eqnarray}

@@

In order to find the hidden state at day $T$, we can simply run the Viterbi
algorithm that gives us the most likely hidden path for the given data and given
HMM. Then, we simply pick the last element from this list, roll the dice and
apply transition once, and this gives us our first prediction for day $T+1$. The
rest will be the same as generating data from a regular HMM - we keep following
the transition table. The pseucode for the function
\verb!observationGeneration!  looks as follows:

@@

\begin{figure}[ht]
\center{
\begin{tabbing}
{\tt next}\={\tt Sta}\={\tt te}$(stat$\=$e,A,\pi,C,R);$\\ 
\> $randVal = ${\tt generateUniform(0,1);}\\
\> $total = 0;$\\
\> $done = false;$\\
\> $total = A[currentState][0];$\\
\> {\tt for } $j = 1$ {\tt to } $numStates$\\
\> \> {\tt if} $done = true$ {\tt break};\\
\> \> {\tt if} $(randVal > total$ {\tt and } \\
\> \> \> $randVal \leq (total + A[currentState][j]))$\\
\> \> \> $currentState = j;$\\
\> \> \> $done = true;$\\
\> \> {\tt end}\\
\> \> $total = total+ A[currentState][j]$;\\
\> {\tt end}\\
\> {\tt return} $currentState$;
\end{tabbing}
}
\caption{State Transition Algorithm}
\vspace{0.6cm}
\end{figure}
